Modal Abstractions for Virtualizing Memory Addresses

### 2025-08-17 relwork.tex
add keywords and CCS; remove all uses of \del

\paragraph{OS Kernel Verification}
There has been relatively little prior work on formal verification of virtual memory.
\replace{Instead, m}Most OS verification work \replace{has focused on minimizing}{minimizes} reasoning about virtual memory management.
The original \textsc{Verisoft} project~\cite{alkassar2008verisoft,alkassar2010pervasive,alkassar2008formal,dalinger2005verification,hillebrand2005address,alkassar2008formal,starostin2010formal} 
relied on custom hardware\del{,} which\del{, among other things,} always ran kernel code with virtual memory disabled, removing the circularity that is a key challenge of verifying
VMM code for real hardware: \replace{A}{a}t that point page tables become a basic partial map data structure to represent user program address translations,
with an idiosyncratic format. \replace{It turns out that s}{S}ubsequent OS verification work
\add{targets real hardware that uses virtual addressing in the kernel, but unsoundly
uses hardware models that do not.} \del{also treats page tables this way, but
unsoundly given that \replace{other projects}{their} target hardware \del{that} \emph{does} run the kernel with address translation:}
\add{Thus} they \emph{trust}
that the particular page table manipulations do not, for example, unmap kernel code
 (which can crash the machine even if done ``temporarily'').\footnote{\add{A ``fun'' bug one author encountered in an industrial prototype kernel.}}
% \footnote{A bug
% one author has personally encountered in a research kernel, though not one discussed here.}).
This is true for \textsc{seL4}~\cite{Klein2009seL4,seL4TOCS,Sewell2013translation}, whose formal machine model omits address translation,
and \textsc{CertiKOS}~\cite{gu15,gu2016certikos,gu2018certikos,chen2016interrupts}, whose refinement proofs rely on
\textsc{CompCert}'s usermode-oriented memory model~\cite{leroy2008formal,leroy2009formally} which assumes
updates to one memory address are independent of updates to another ---
which is not true of page table updates.
Other work on OS verification either never progressed \del{far enough} to \del{address} VMM verification
(\textsc{Verisoft XT}~\cite{cohen2009vcc,cohen2010local,dahlweid2009vcc,cohen2013SOFSEM}),
or uses memory-safe languages
\replace{to enable safe cohabitation of a single address space by all processes}{for process isolation instead of address translation}
(\textsc{Singularity}~\cite{Fahndrich2006language,Hunt2007singularity,Hunt2007sealing,Barnett2011specsharp}, \textsc{Verve}~\cite{Yang2010Verve},
and \textsc{Tock}~\cite{levy2017multiprogramming})\replace{. The latter approach ensures}{, ensuring} memory safety, but ignor\replace{es}{ing} other functional uses of virtual memory
hardware, like swapping~\cite{Denning1970VM} or exploiting copy-on-write techniques for dynamic migration of virtual machines~\cite{clark2005live}.
\looseness=-1

@@

\replace{Unlike our work, Kolanski and Klein prove very useful embedding theorems stating that code that does not modify page table 
entries can be verified in a VM-ignorant program logic, and that proofs in that logic can be embedded into the VM-aware logic 
(essentially by interpreting ``normal'' points-to relations as virtual points-to facts).
}{
 Kolanski and Klein prove that updates to any memory locations that are not part of the page tables
 do not affect the interpretation of other memory addresses, just like on real hardware.
 This implies that programs that do not modify memory mappings could be reasoned about without
 concern for mappings.
}
\replace{While we have not proven {such a result},
an analagous result \replace{{should} hold}{holds} of our work:}{
 An analagous result should holds of our model (though we have not proven it).
}
\replace{ consider that the doubles for the}{Informally this is visible in the rules for}
\texttt{mov} instructions,
\replace{that access memory behave just as one would expect for}{which are nearly identical to rules in a} VM-ignorant logic~\cite{Chlipala2013Bedrock,ni2007contexts}.
\replace{With our general approach to virtual points-to assertions being inspired by Kolanski and Klein, \emph{both}
 our approach and theirs could in principle}{
  In principle both our approach and Kolanski and Klein's could
 } be extended to account for pageable points-to assertions by adding \del{additional}
disjunctions to an extended points-to definition\add{,}\del{; embedding ``regular'' separation logic into such a variant
is} \add{which would be}
 the appropriate next step to extend reasoning to usermode programs running with a kernel that may demand-page the program's
memory.
\looseness=-1

% \todo[inline]{author response talks about syeda and klein here briefly, and it fits, but we also want the longer discussion}

\del{Translation Lookaside Buffers}
As noted in Section \ref{sec:backgroundonmachinemodel}, we do not formally model or reason about \add{translation lookaside buffers} \add{(}TLBs\add{)}.
\replace{TLB flushes are required only when addresses are removed from a virtual address space, or when changing
virtual address spaces.}{
 TLB flushes are necessary when a page is \emph{un}mapped, or when switching address spaces.
}
\replace{Because t}{T}his occurs in few places in \replace{the}{uniprocessor} kernel\add{s} (in some, only 3 locations), \add{so} full verified
kernels including \textsc{seL4}~\cite{Klein2009seL4,seL4TOCS} and \textsc{CertiKOS}~\cite{gu15,gu2016certikos}
trust TLB management. Neither of the aforementioned systems has a hardware model including a TLB, so neither is able
to verify TLB management in any form --- they \emph{must} trust its operation.
\replace{
The only place TLB code becomes particularly challenging is in multiprocessor kernels, where
not only must the running CPU flush its own TLB with special instructions, but it must send an inter-processor interrupt (IPI) to all
other cores to ensure they also flush the relevant ranges of their own TLBs.
Currently no formal hardware model exists with sufficient detail to reason about IPIs in full detail,
which would require extensions to model interactions with ACPI tables populated by firmware on desktop and server machines,
or flattened device trees for (semi-)embedded systems like single-board ARM and RISC-V machines, which tell the OS
important information about how to control other CPU cores;
plus the memory-mapped IO used to trigger those IPIs.
As a result, multicore extensions of \textsc{seL4}~[?] and \textsc{CertiKOS}~[?] also trust this functionality.
}{
 This is true of multicore verified kernels as well~\cite{von2013clustered,gu2016certikos}, though there
 the situation becomes much more challenging: when unmapping pages, the running CPU must invalidate the relevant TLB entries locally,
 but also send an inter-procesor interrupt (IPI) to all other cores to ensure they also invalidate affected entries on their TLBs.
 No formal hardware model currently exists for IPIs on any architecture, or even for the memory-mapped IO
 used to trigger those IPIs.
}

The only \replace{other attempt to verify a change of address space was by}{work to tackle TLB code verification was}
\citet{syeda2018program,syeda2020formal},
\replace{whose focus was on TLB maintenance (which we do not model)}{who are also the only prior work on verifying address-space-aware context switching}.
However, they verified only the 4 instructions to switch address spaces and update the TLB on an ARM processor, in isolation
(i.e., not the full context switch including changing stacks with address spaces).
The specification they proved for those instructions did not address program invariants that may be valid in one address space and not the other,
so is not flexible enough to extend directly to a full context switching primitive as in Figure \ref{fig:swtchC}\replace{,}{.}
\del{in addition to inheriting other limitations from {the work they extend, discussed above}.}
\add{Their logic adds an assertion tracking known-inconsistent addresses (i.e., recently-unmapped by a table update or change in page table root)
and their memory access rules require accessed memory to lie outside that set.
However their logic is proven sound against a generalized semantics where any page table change adds \emph{all}
addresses to the inconsistent set, which is too imprecise for finer-grained unmapping.
}
The right \replace{eventual}{general} solution would be to combine our work
and \add{an extension of} theirs, \add{a substantial project} which we leave to future work\replace{,}{.}\del{ as combining both the models and reasoning principles would be highly non-trivial.}
\add{
 No other prior work has considered address space changes during context switching. \textsc{XCAP}~\cite{ni2007contexts}
 and Bedrock~\cite{Chlipala2011Bedrock,Chlipala2013Bedrock,Chlipala2015webapp} deal
 only with usermode threading (in a single process). \textsc{CertiKOS} and \textsc{seL4} trust assembly primitives for
 context switches, and do  not model address translation for executing code.
 \looseness=-1
}

>>>>

\paragraph{OS Kernel Verification}
There has been relatively little prior work on formal verification of virtual memory.
\replace{Instead, m}Most OS verification work \replace{has focused on minimizing}{minimizes} reasoning about virtual memory management.
The original \textsc{Verisoft} project~\cite{alkassar2008verisoft,alkassar2010pervasive,alkassar2008formal,dalinger2005verification,hillebrand2005address,alkassar2008formal,starostin2010formal} 
relied on custom hardware which always ran kernel code with virtual memory disabled, removing the circularity that is a key challenge of verifying
VMM code for real hardware: \replace{A}{a}t that point page tables become a basic partial map data structure to represent user program address translations,
with an idiosyncratic format. \replace{It turns out that s}{S}ubsequent OS verification work
\add{targets real hardware that uses virtual addressing in the kernel, but unsoundly
uses hardware models that do not.} 
\add{Thus} they \emph{trust}
that the particular page table manipulations do not, for example, unmap kernel code
 (which can crash the machine even if done ``temporarily'').\footnote{\add{A ``fun'' bug one author encountered in an industrial prototype kernel.}}
% \footnote{A bug
% one author has personally encountered in a research kernel, though not one discussed here.}).
This is true for \textsc{seL4}~\cite{Klein2009seL4,seL4TOCS,Sewell2013translation}, whose formal machine model omits address translation,
and \textsc{CertiKOS}~\cite{gu15,gu2016certikos,gu2018certikos,chen2016interrupts}, whose refinement proofs rely on
\textsc{CompCert}'s usermode-oriented memory model~\cite{leroy2008formal,leroy2009formally} which assumes
updates to one memory address are independent of updates to another ---
which is not true of page table updates.
Other work on OS verification either never progressed to VMM verification
(\textsc{Verisoft XT}~\cite{cohen2009vcc,cohen2010local,dahlweid2009vcc,cohen2013SOFSEM}),
or uses memory-safe languages
\replace{to enable safe cohabitation of a single address space by all processes}{for process isolation instead of address translation}
(\textsc{Singularity}~\cite{Fahndrich2006language,Hunt2007singularity,Hunt2007sealing,Barnett2011specsharp}, \textsc{Verve}~\cite{Yang2010Verve},
and \textsc{Tock}~\cite{levy2017multiprogramming})\replace{. The latter approach ensures}{, ensuring} memory safety, but ignor\replace{es}{ing} other functional uses of virtual memory
hardware, like swapping~\cite{Denning1970VM} or exploiting copy-on-write techniques for dynamic migration of virtual machines~\cite{clark2005live}.
\looseness=-1

@@

\replace{Unlike our work, Kolanski and Klein prove very useful embedding theorems stating that code that does not modify page table 
entries can be verified in a VM-ignorant program logic, and that proofs in that logic can be embedded into the VM-aware logic 
(essentially by interpreting ``normal'' points-to relations as virtual points-to facts).
}{
 Kolanski and Klein prove that updates to any memory locations that are not part of the page tables
 do not affect the interpretation of other memory addresses, just like on real hardware.
 This implies that programs that do not modify memory mappings could be reasoned about without
 concern for mappings.
}
\replace{While we have not proven {such a result},
an analagous result \replace{{should} hold}{holds} of our work:}{
 An analagous result should holds of our model (though we have not proven it).
}
\replace{ consider that the doubles for the}{Informally this is visible in the rules for}
\texttt{mov} instructions,
\replace{that access memory behave just as one would expect for}{which are nearly identical to rules in a} VM-ignorant logic~\cite{Chlipala2013Bedrock,ni2007contexts}.
\replace{With our general approach to virtual points-to assertions being inspired by Kolanski and Klein, \emph{both}
 our approach and theirs could in principle}{
  In principle both our approach and Kolanski and Klein's could
 } be extended to account for pageable points-to assertions by adding 
disjunctions to an extended points-to definition\add{,} \add{which would be}
 the appropriate next step to extend reasoning to usermode programs running with a kernel that may demand-page the program's
memory.
\looseness=-1

As noted in Section \ref{sec:backgroundonmachinemodel}, we do not formally model or reason about \add{translation lookaside buffers} \add{(}TLBs\add{)}.
\replace{TLB flushes are required only when addresses are removed from a virtual address space, or when changing
virtual address spaces.}{
 TLB flushes are necessary when a page is \emph{un}mapped, or when switching address spaces.
}
\replace{Because t}{T}his occurs in few places in \replace{the}{uniprocessor} kernel\add{s} (in some, only 3 locations), \add{so} full verified
kernels including \textsc{seL4}~\cite{Klein2009seL4,seL4TOCS} and \textsc{CertiKOS}~\cite{gu15,gu2016certikos}
trust TLB management. Neither of the aforementioned systems has a hardware model including a TLB, so neither is able
to verify TLB management in any form --- they \emph{must} trust its operation.
\replace{
The only place TLB code becomes particularly challenging is in multiprocessor kernels, where
not only must the running CPU flush its own TLB with special instructions, but it must send an inter-processor interrupt (IPI) to all
other cores to ensure they also flush the relevant ranges of their own TLBs.
Currently no formal hardware model exists with sufficient detail to reason about IPIs in full detail,
which would require extensions to model interactions with ACPI tables populated by firmware on desktop and server machines,
or flattened device trees for (semi-)embedded systems like single-board ARM and RISC-V machines, which tell the OS
important information about how to control other CPU cores;
plus the memory-mapped IO used to trigger those IPIs.
As a result, multicore extensions of \textsc{seL4}~[?] and \textsc{CertiKOS}~[?] also trust this functionality.
}{
 This is true of multicore verified kernels as well~\cite{von2013clustered,gu2016certikos}, though there
 the situation becomes much more challenging: when unmapping pages, the running CPU must invalidate the relevant TLB entries locally,
 but also send an inter-procesor interrupt (IPI) to all other cores to ensure they also invalidate affected entries on their TLBs.
 No formal hardware model currently exists for IPIs on any architecture, or even for the memory-mapped IO
 used to trigger those IPIs.
}

The only \replace{other attempt to verify a change of address space was by}{work to tackle TLB code verification was}
\citet{syeda2018program,syeda2020formal},
\replace{whose focus was on TLB maintenance (which we do not model)}{who are also the only prior work on verifying address-space-aware context switching}.
However, they verified only the 4 instructions to switch address spaces and update the TLB on an ARM processor, in isolation
(i.e., not the full context switch including changing stacks with address spaces).
The specification they proved for those instructions did not address program invariants that may be valid in one address space and not the other,
so is not flexible enough to extend directly to a full context switching primitive as in Figure \ref{fig:swtchC}\replace{,}{.}
\add{Their logic adds an assertion tracking known-inconsistent addresses (i.e., recently-unmapped by a table update or change in page table root)
and their memory access rules require accessed memory to lie outside that set.
However their logic is proven sound against a generalized semantics where any page table change adds \emph{all}
addresses to the inconsistent set, which is too imprecise for finer-grained unmapping.
}
The right \replace{eventual}{general} solution would be to combine our work
and \add{an extension of} theirs, \add{a substantial project} which we leave to future work\replace{,}{.}
\add{
 No other prior work has considered address space changes during context switching. \textsc{XCAP}~\cite{ni2007contexts}
 and Bedrock~\cite{Chlipala2011Bedrock,Chlipala2013Bedrock,Chlipala2015webapp} deal
 only with usermode threading (in a single process). \textsc{CertiKOS} and \textsc{seL4} trust assembly primitives for
 context switches, and do  not model address translation for executing code.
 \looseness=-1
}
